---
title: "Reproducible Research: Peer Assessment 1"
author: "John Southgate"
date: "12th January 2017"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Investigate data from a personal activity monitoring device

## Setup
```{r}
library(dplyr)
```

```{r echo=FALSE}
setwd("C:/Users/go2/_my/CourseraDataSci/RepData_PeerAssessment1")
#getwd()
#dir()
```

## Loading and preprocessing the data
```{r}
activity <- read.csv("activity.csv", colClasses = "character")
str(activity)
head(activity)
summary(activity$steps)
table(activity$steps)
### NB: 11014 inactive intervals - no steps
```

## Pre-process the table for the correct formats
```{r}
activity$nsteps<-as.numeric(activity$steps)
activity$Date<-as.Date(activity$date)
activity$Interval<-as.numeric(activity$interval)
```

## What is mean total number of steps taken per day?
### Calculate the total number of steps taken per day
Ignoring days with no data
```{r}
daily.total<-aggregate(nsteps ~ Date, sum, data=activity,na.rm=TRUE)
head(daily.total)
```
check
```{r}
daily.steps<-split(activity$nsteps, activity$Date)
head(daily.steps)
```

### Make a histogram of the total number of steps taken each day
include inactive intervals
```{r}
plot(daily.total$Date,daily.total$nsteps,type="h", main="Total steps taken each day")
```

## Calculate and report the mean and median of the steps taken per day
mean: include inactive intervals
```{r}
daily.mean<-aggregate(nsteps ~ Date, mean, data=activity,na.rm=TRUE)
head(daily.mean)
```

median: include inactive intervals
```{r}
daily.median<-aggregate(nsteps ~ Date, median, data=activity,na.rm=TRUE)
head(daily.median)
```

## What is the average daily activity pattern?
### Calculate average number of steps taken, averaged across all days (y-axis)
```{r}
interval.mean<-aggregate(nsteps ~ Interval, mean, data=activity,na.rm=TRUE)
head(interval.mean)
plot(interval.mean$Interval,interval.mean$nsteps,type="l", main="Mean steps across all days")
```

## Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
Now use the dplyr objects (tibble) and functions 
```{r}
df<-tbl_df(activity)
#class(df)
Intervals<-group_by(df, Interval)
#Intervals %>% print(n = nrow(.))
Intervals.max.steps <- summarize(Intervals, max.steps = max(nsteps, na.rm=T))
#Intervals.max.steps %>% print(n = nrow(.))
Interval.max<-arrange(Intervals.max.steps, desc(max.steps))[1,]$Interval
#Interval.max
print (paste("Maximum steps occurs in the",Interval.max,"minute interval"))
```

## Imputing missing values

### Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
activity.cases<-complete.cases(activity) 
ac<-activity[activity.cases,]
print(paste((nrow(activity)-nrow(ac)),"missing rows"))
```

### Devise a strategy for filling in all of the missing values in the dataset. 
Just take the mean of steps for interval across all days and apply it to each interval with missing values 
```{r}
Intervals.mean.steps <- round(summarize(Intervals, mean.steps = mean(nsteps, na.rm=T)))
#Intervals.mean.steps %>% print(n = nrow(.))
#nrow(Intervals.mean.steps)
```

### Create a new dataset that is equal to the original dataset but with the missing data filled in
```{r}
activity2<-activity

for (i in 1:nrow(activity)) 
{
	if (is.na(activity$nsteps[i])) 
	{activity2$nsteps[i] <- Intervals.mean.steps[(Intervals.mean.steps$Interval==activity$Interval[i]),]$mean.steps
	}
	else
	{activity2$nsteps[i] <- activity$nsteps[i]
	}
	activity2$steps[i] <- as.character(activity2$nsteps[i])
	}
head(activity2,20)
head(Intervals.mean.steps,20)
```

### Calculate the total number of steps taken per day
```{r}
daily2.total<-aggregate(nsteps ~ Date, sum, data=activity2,na.rm=TRUE)
head(daily2.total)
```
check
```{r}
daily2.steps<-split(activity2$nsteps, activity2$Date)
head(daily2.steps)
```

### Make a histogram of the total number of steps taken each day
include inactive intervals
```{r}
plot(daily2.total$Date,daily2.total$nsteps,type="h", main="Total steps for each day, with imputed values")
```

## Calculate and report the mean and median of the steps taken per day
mean: include inactive intervals
```{r}
daily2.mean<-aggregate(nsteps ~ Date, mean, data=activity2,na.rm=TRUE)
```

median: include inactive intervals
```{r}
daily2.median<-aggregate(nsteps ~ Date, median, data=activity2,na.rm=TRUE)
head(daily.mean)
head(daily2.mean)
head(daily.median)
head(daily2.median)

par(mfrow=c(2,1))
plot(daily.total$Date,daily.total$nsteps,type="h", main="Total steps taken each day with missing values")
lines(daily.total$Date,daily.total$nsteps,type="l", col="blue", lwd=2)
plot(daily2.total$Date,daily2.total$nsteps,type="h", main = "Total steps taken each day, no missing values")
lines(daily2.total$Date,daily2.total$nsteps,type="l", col="blue", lwd=2)
```

### Do these values differ? What is the impact ?
These values differ. Missing histogram bars are filled in because more dates have values. Smoothing the histogram makes significant trends easier to spot. 

### Plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days 

Populate daytype
```{r}
activity2["daytype"]<-NA
for (i in 1:nrow(activity2)) {
if (weekdays(activity2$Date[i])  %in%  c("Saturday","Sunday"))
   {activity2$daytype[i]<-"weekend"}
else
   {activity2$daytype[i]<-"weekday"}
}

#unique(activity2$daytype)
```

Filter and plot by daytype
```{r}
par(mfrow=c(1,1))
df_weekday=filter(activity2, daytype == "weekday")
df_weekend=filter(activity2, daytype == "weekend")

interval2.mean.wd<-aggregate(nsteps ~ Interval, mean, data=df_weekday,na.rm=TRUE)
interval2.mean.wd2<-mean(df_weekday$nsteps)
head(interval2.mean.wd)
plot(interval2.mean.wd$Interval,interval2.mean.wd$nsteps,type="l",col="red", lwd=1, main="Comparison of Steps: Weekdays and Weekends")

interval2.mean.we<-aggregate(nsteps ~ Interval, mean, data=df_weekend,na.rm=TRUE)
interval2.mean.we2<-mean(df_weekend$nsteps)
head(interval2.mean.we)
lines(interval2.mean.we$Interval,interval2.mean.we$nsteps,type="l",col="blue", lwd=1)

abline(h=interval2.mean.wd2,lty=2,col="red")
abline(h=interval2.mean.we2,lty=2,col="blue")
legend("topright",lty=1, col=c("red","blue"),c("weekdays","weekends"))
```

Greater number of steps are observed on weekdays in the lower 5 minute intervals. From about the 900 minute interval, greater steps are observed on weekends. As can be seen by the dashed lines, more steps are taken on average on weekends.
